Robust Bayesian Analysis of Causal Inference in Medical Diagnosis and Treatment

### 2024-07-31 JASA-template.tex
Fix notation, add sentence to clarify strength of probit assumption.

To decide whether or not to treat a new individual with given predictors,
we are mainly interested in the effect of the treatment on the outcome.
More precisely, the causal effect of a new individual, indexed as $n+1$,
whose outcome $Y_{n+1}$ is not yet observed, and with observed predictors $X_{n+1}=x_{n+1}$, is defined by:
\begin{align}
  \begin{split}
  \delta(x_{n+1})
    &\coloneqq\mathbb{E}(Y_{n+1}\mid X_{n+1}=x_{n+1},T_{n+1} =1) \\
    &\qquad - \mathbb{E}(Y_{n+1}\mid X_{n+1}=x_{n+1},T_{n+1}=0)
  \end{split}
  \intertext{For our model, due to linearity of expectation, we have that}
  \begin{split}
  \delta(x_{n+1})
    &=\beta_T
      +x_{n+1}\beta+\mathbb{E}(\epsilon_{n+1}\mid X_{n+1}=x_{n+1},T_{n+1} =1) \\
    &\qquad\,\,\, %%% for alignment
      -x_{n+1}\beta-\mathbb{E}(\epsilon_{n+1}\mid X_{n+1}=x_{n+1},T_{n+1} =0)
  \end{split}
\intertext{and because $\epsilon_{n+1}$ is independent from $X_{n+1}$ and $T_{n+1}$,}
  &=\beta_{T}.
\end{align}
Note that, for this model, the causal effect $\delta(x_{n+1})$
does not depend on the observed value $x_{n+1}$ of $X_{n+1}$.
So, to find the causal effect, we simply need to estimate $\beta_T$.
\added{%
Note that, if interaction terms between $X_i$ and $T_i$ were present in
the model (for example a term of the form, say, $\eta X_iT_i$ for some
parameter vector $\eta$), that would result in a dependence of the
causal effect on $x_{n+1}$.%
}

To estimate $\beta_T$ from the data $X$, $Y$ and $T$,
especially in the presence of confounders,
we also need to consider the
association between the treatment indicators $T$ and the predictors $X$.
A common choice in the literature is to use a probit link function
\citep{winship99},
\added{%
though other link functions, such as the logit, can also be used
\citep{HECKMAN1985}.%
}
In this way, we can
specify the conditional probability that subject $i$ receives the treatment through a linear model. 
That is, for another vector of regression coefficients 
$\gamma\coloneqq(\gamma_1, \cdots, \gamma_p)^{\top}$ we
assume
\begin{align}
	P(T_i=1\mid X_i) = \Phi(X_i\gamma)
\end{align}
where $\Phi$ denotes the cumulative distribution function
of a standard normal distribution. To incorporate this probit
link function, we model the $T_i$ as follows \citep{albert93}:
\begin{align}
    T_i^* &= X_i\gamma +u_i \\
    T_i   &= \mathbb{I}(T_i^*>0)
    =
    \begin{cases}
    1 & \text{if }T_i^*>0 \\
    0 & \text{otherwise}
    \end{cases}
\end{align}
where $u_i\sim\mathcal{N}(0,1)$.
With this model, indeed
\begin{align}
  P(T_i=1\mid X_i)
  &=P(T_i^*>0)=P(u_i>-X_i\gamma)=1-P(u_i\le -X_i\gamma) \\
  &=1-\Phi(-X_i\gamma)=\Phi(X_i\gamma).
\end{align}

>>>>

To decide whether or not to treat a new individual with given predictors,
we are mainly interested in the effect of the treatment on the outcome.
More precisely, the causal effect of a new individual, indexed as $n+1$,
whose outcome $Y_{n+1}$ is not yet observed, and with observed predictors $X_{n+1}=x_{n+1}$, is defined by:
\begin{align}
  \begin{split}
  \delta(x_{n+1})
    &\coloneqq\mathbb{E}(Y_{n+1}\mid X_{n+1}=x_{n+1},T_{n+1} =1) \\
    &\qquad - \mathbb{E}(Y_{n+1}\mid X_{n+1}=x_{n+1},T_{n+1}=0)
  \end{split}
  \intertext{For our model, due to linearity of expectation, we have that}
  \begin{split}
  \delta(x_{n+1})
    &=\beta_T
      +x_{n+1}\beta+\mathbb{E}(\epsilon_{n+1}\mid X_{n+1}=x_{n+1},T_{n+1} =1) \\
    &\qquad\,\,\, %%% for alignment
      -x_{n+1}\beta-\mathbb{E}(\epsilon_{n+1}\mid X_{n+1}=x_{n+1},T_{n+1} =0)
  \end{split}
\intertext{and because $\epsilon_{n+1}$ is independent from $X_{n+1}$ and $T_{n+1}$,}
  &=\beta_{T}.
\end{align}
Note that, for this model, the causal effect $\delta(x_{n+1})$
does not depend on the observed value $x_{n+1}$ of $X_{n+1}$.
So, to find the causal effect, we simply need to estimate $\beta_T$.
\added{%
Note that, if interaction terms between $X_i$ and $T_i$ were present in
the model (for example a term of the form, say, $T_iX_i\eta$ for some
parameter vector $\eta$), that would result in a dependence of the
causal effect on $x_{n+1}$.%
}

To estimate $\beta_T$ from the data $X$, $Y$ and $T$,
especially in the presence of confounders,
we also need to consider the
association between the treatment indicators $T$ and the predictors $X$.
A common choice in the literature is to use a probit link function
\citep{winship99},
\added{%
though other link functions, such as the logit, can also be used
\citep{HECKMAN1985}.%
}
In this way, we can
specify the conditional probability that subject $i$ receives the treatment through a linear model. 
That is, for another vector of regression coefficients 
$\gamma\coloneqq(\gamma_1, \cdots, \gamma_p)^{\top}$ we
assume
\begin{align}
	P(T_i=1\mid X_i) = \Phi(X_i\gamma)
\end{align}
where $\Phi$ denotes the cumulative distribution function
of a standard normal distribution.
\added{%
The key assumption made here is that
there is a monotone relationship between
the predictors and the probability of treatment.
Here too, interaction terms between the $X_i$ could be added
to form more complex models if so desired.%
}

To incorporate this probit
link function, we model the $T_i$ as follows \citep{albert93}:
\begin{align}
    T_i^* &= X_i\gamma +u_i \\
    T_i   &= \mathbb{I}(T_i^*>0)
    =
    \begin{cases}
    1 & \text{if }T_i^*>0 \\
    0 & \text{otherwise}
    \end{cases}
\end{align}
where $u_i\sim\mathcal{N}(0,1)$.
With this model, indeed
\begin{align}
  P(T_i=1\mid X_i)
  &=P(T_i^*>0)=P(u_i>-X_i\gamma)=1-P(u_i\le -X_i\gamma) \\
  &=1-\Phi(-X_i\gamma)=\Phi(X_i\gamma).
\end{align}
